---
title: "Humanities majors yikes"
author: "Ben Schmidt"
date: "August 8, 2018"
output:
  word_document: default
  html_document: default
---

The crisis of the humanities is not news.
Whil coverage of individual academic disciplines like musicology, history, or comparative literature often deals with the substance of scholarship, 
talk of the humanities in general *always* seems to focus on their imminent extinction.
That long history means conversations about the fate of the humanities can track the same rut for a lifetime, absent any evidence to ground them.
[A bestselling volume in 1964](https://catalog.hathitrust.org/Record/001450049) warned that a science-focused world left no room for humane pursuits, even as baby boomers began to flood the English and history departments of new universities. Allan Bloom warned about academics putting liberal ideology before scholarship in 1987: humanities degrees quickly rose.
In 2010, Wayne Bivens-Tatum provided a [useful walk through the first 50 years of the humanities crisis, until about 1980](https://blogs.princeton.edu/librarian/2010/11/the_crisis_in_the_humanities/).


```{r, include = FALSE}
knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE, cache = TRUE, fig.width = 8, fig.height = 4)
```

```{r, cache = TRUE}
# Preparatory functions to get data.
source("read_cips.R")

characteristics = read_csv("hd2017.csv")
cips = return_cips()
data = return_data()

source("EDA_functions.R")

data = data %>%
  filter(CIPCODE != "99.0000", CIPCODE != "99") %>%
  mutate(AWLEVEL=as.numeric(AWLEVEL),MAJORNUM=as.numeric(MAJORNUM))

totals = data %>%
  filter(AWLEVEL==5) %>%
  group_by(year) %>% 
  summarize(year_total = sum(degrees, na.rm=T), people_total = sum(degrees[MAJORNUM==1], na.rm=T))

long_term = read_tsv("long_term.txt")
long_term2 = read_csv("long_term_gender.csv") %>% gather(variable, degrees, -year) %>% separate(variable, c("gender","Discipline")) %>% filter(year < 2000) %>% mutate(gender = ifelse(gender=="Men", "men", "women"))

gender_totals = long_term2 %>% filter(Discipline=="All") %>% select(-Discipline) %>% spread(gender, degrees) %>% 
  bind_rows(data %>% filter(AWLEVEL==5, MAJORNUM==1, year > 1999) %>% group_by(year, gender) %>% summarize(degrees = sum(degrees)) %>%  ungroup %>% spread(gender, degrees))

ltotals = totals %>% group_by(year) %>% summarize(year_total = sum(year_total)) %>% mutate(source="IPEDS") %>% bind_rows(
  long_term %>% mutate(source="Old", year_total = total) %>% select(year=Year, year_total, people_total=total, source )
  ) %>% left_join(gender_totals)

pop = read_csv("pop_estimates.csv") %>% select(-junk)

tot_degrees = ltotals %>% group_by(year) %>% summarize(total_degrees = max(year_total))

```

Because of this long history, I've always been skeptical of claims that the humanities are in retreat. But something different has been happening with the humanities since the 2008 financial crisis. Five years ago, I was still arguing that the humanities were [still near long-term norms in their number of majors.](http://www.chronicle.com/blognetwork/edgeofthewest/2013/06/10/the-humanities-crisis/)
But since then, I've been watching the [major numbers from the Department of Education](https://nces.ed.gov/ipeds/),
and every year things look worse. Almost every humanities field has seen a rapid drop in majors: history is down about 45% from its 2007 peak, 
while English has nearly halved since the late 1990s.
Student majors have dropped, rapidly, at a variety of types of institutions. Declines have hit almost every field in the humanities (with one interesting exception) and related social sciences, they have not stabilized with the economic recovery, and they appear to reflect a new set of student priorities, which are being formed even before they see the inside of a college classroom.

```{r, fig.height=6, fig.width=7}
plottable = data %>% filter(MAJORNUM==1, AWLEVEL==5) %>% 
  into_major_nums %>% 
  left_join(cips,by=c("CIPCODE"="CIPCode"))

this_plot = plottable %>% filter(year >= 1997) %>%
    group_by(year,General,Type) %>%
    filter(General %in% c("Humanities")) %>% 
    filter(Type %in% c("English","History","Languages and Literature","Philosophy", "Classical Studies","Religion")) %>%
    summarize(degrees = sum(degrees), percent = sum(degrees)/year_total[1]) %>% filter(!(Type=="Area Studies" && year < 1997)) %>%
    group_by(Type) %>%
    mutate(relative=percent/max(percent))

this_plot %>% ggplot(aes(x=year,y=relative-1,fill=Type,lty=Type,label=Type,color=Type)) +
    geom_line() + 
    scale_y_continuous("Share relative to peak", labels=scales::percent) + 
#    facet_wrap(~Type,scales="free_y") + 
   theme(legend.position="none") + 
    labs(title="The big humanities majors\nhave dropped rapidly in the last decade to decline",subtitle="As percentage of all BAs; all US institutions",
         caption = "IPEDS data (preliminary for 2017)\nChart by Ben Schmidt, 2018") + 
  geom_point(data = this_plot %>% filter(relative==1)) + 
  geom_text(data=this_plot %>% filter(year==2017), adj=0, nudge_x = .1) + theme_bw() + theme(legend.position = "none") + expand_limits(x=2026)

```

One thing I learned earning a history degree is that people usually announce a "crisis" so they can trot out solutions they came up with years earlier.
I don't have any right now. But the drop in majors since 2008 has been so intense that I now think there is, 
in the only meaningful sense of the word, a crisis. That is: we are in a momentum of rapid change in which decisions are 
especially important, and will have continuing ramifications for what American universities look like.
But right now, the biggest impediment to thinking about the future of the humanities is that, thanks to this entrenched narrative of decline—because we’ve been crying wolf for so long—we already think we know what’s going on.

### The two crises of the humanities

The most reliable indicators about the humanities in American colleges are reports that all colleges and universities make to the Department of Education. These time series run back to about 1950. Since then, the humanities have seen three eras. The first ran from 1955 to 1985. As normal schools around the country transformed into comprehensive universities, men and women alike poured into English and History majors; then, when the economy soured and the growth of higher education slowed in the 1970s, the boom turned to bust, and humanities majors collapsed nationwide. The second phase ran began around 1985 and ran to 2008. This was a long period of stability; majors in the four largest (and easiest to track over the long term) humanities majors held steady, with modest fluctuations. Since 2008, the crisis of the humanities has resumed, with percentage drops that are beginning to approach those of 40 years ago. Unlike the drops of the 70s, though, there's no pre-existing bubble to deflate. Nor is there a compelling demographic explanation. Five years ago, it was reasonable to look at these numbers and conclude [the long-term story is all about gender](https://www.theatlantic.com/education/archive/2013/12/the-real-reason-the-humanities-are-in-crisis/282441/). Men majored in humanities fields at the same rate in the 1990s as they had in the 1950s, while women, seeing more options in the workforce, increasingly turned to majors in business fields. But the drops since the financial crisis have been in men and women, across racial groups, and spread across a wide variety of universities.

```{r}
majornums = data %>% 
  filter(AWLEVEL==5) %>%
  group_by(year, CIPCODE, MAJORNUM, gender) %>%
  summarize(degrees = sum(degrees)) %>%
  acadjoin()

new_series = majornums %>%
  group_by(year, Discipline, gender) %>%
  filter(Bachelors=="Humanities", MAJORNUM==1) %>%
  summarize(degrees = sum(degrees, na.rm=T)) %>%
  ungroup %>% 
  filter(!is.na(Discipline)) %>%
  group_by(Discipline, year, gender) %>%
  summarize(degrees = sum(degrees, na.rm=T)) %>%
  ungroup %>% 
  inner_join(totals) %>% 
  mutate(percent = degrees/year_total) %>%
  group_by(Discipline) %>% 
  group_by(Discipline) %>%
  mutate(size = factor(round(.5 + log(mean(percent), base = 9))))

joint_series = long_term2 %>% mutate(source = "old") %>% select(year, Discipline, gender, degrees = degrees, source) %>% 
  filter(Discipline != "All", year < 2000) %>% 
  bind_rows(
    new_series %>% 
      #filter(grepl("(English Language and Literature|History|Languages|Philosophy)", Discipline)) %>% 
      filter(!grepl("Liberal Studies)", Discipline)) %>%
      select(year, Discipline, degrees, gender) %>% mutate(source="new") %>% filter(year >= 2000)
  ) %>% 
  mutate(Discipline = fct_recode(Discipline, c("English" = "English Language and Literature"), c("Languages" = "Languages and Literatures Other than English"))) %>%
  inner_join(pop) %>%
  inner_join(tot_degrees) %>%
  group_by(Discipline, gender) %>% 
  mutate(years = sum(degrees)) %>%
  mutate(longterm = years > 300000) %>%
  distinct(year, Discipline, gender, .keep_all = T)
```

```{r}
this_plot = joint_series %>% filter(Discipline %in% c("English", "History", "Languages", "Philosophy")) %>% group_by(year, total_degrees, gender) %>% summarize(humanities=sum(degrees)) %>% filter(year >= 1950) %>% inner_join(gender_totals %>% gather("gender", "total", men, women))

#this_plot %>% ggplot() + geom_text(data = this_plot %>% filter(year %in% c(1955, 1960, 1964, 1967, 1969, 1971, 1973, 1977, 1984, 1993, 2000, 2006, 2013, 2017))) + geom_path() + aes(x=total, y = humanities, label=year, color=gender) + expand_limits(y = 0, x = 0) + scale_y_continuous("Degrees in English, Philosophy, History, Languages", labels = scales::comma) + scale_x_continuous("Overall BA degrees", labels = scales::comma)

labels = frame_data(
  ~label, ~gender, ~percent, ~year,
  "Women", "women", .2, 1978,
  "Men", "men", .05, 1977
)

this_plot %>% mutate(percent = humanities/total) %>% ggplot() + aes(x=year,  y = percent, color=gender) + geom_line(lwd=1.2) + theme_bw() + scale_y_continuous("History, Philosophy, English, and Languages\nas percent of all degrees", labels=scales::percent) + expand_limits(y=0) + theme(legend.position="none") + labs(title="Long-term Humanities Degree share in American Universities", subtitle="Four largest degree fields") + scale_color_brewer(palette = "Accent", type="qual") + geom_text(data = labels, size=5, aes(label = label))

```

College degrees are a somewhat problematic metric: I'd rather see information about the type and level of courses that undergraduates take. This kind of data is hard to come by except at an anecdotal level. But where large scale enrollment numbers do exist, they are not especially reassuring: the American Historical Association surveyed history departments and found [course enrollments dropping by 7.7% from 2014 to 2017](https://www.historians.org/publications-and-directories/perspectives-on-history/february-2018/enrollment-declines-continue-aha-survey-again-shows-fewer-undergraduates-in-history-courses), with upper level courses seeing greater declines than introductory ones. This is bad, but not as catastrophic as the major numbers, which fell by 20% in the same period. That suggests that either the declines are beginning to stabilize, or that students are more willing to spend elective credits on humanities courses than to major in them.

There's another way that degree 'share' is a bit strange: American higher education continuously expands to include new populations, and you might expect humanities degrees to be the province only of a certain sort of elite institution. And the humanities are multiplying: there weren't many of Chicano Studies majors in  1968. But even seen in this light, the humanities are in retreat. Looking at degrees as a share of the entire American 23-year-old population shows that there is, indeed, some moderate increase nationwide in the extremely long term. But even after adding a raft of fields like ethnic and gender studies, musicology and art history, and religion, a lower share of newly-graduated Americans have humanities degrees than did in 1970 or 1990. Even the absolute *number* is lower than in 1970: the big four fields are flirting with the 100,000 degree line, which they first passed in 1968 and haven't been below for almost twenty years. (If you include mass communication and media studies, which mixes social science, humanities, and professional education together, things look a bit better: but increases in communication majors since 2008 nationwide can only offset a single year's loss of English BAs.)

```{r}
library(ggrepel)
joint_series = majornums %>% filter(MAJORNUM==1) %>%
  group_by(year, Discipline) %>%
  filter(Bachelors=="Humanities") %>%
  summarize(degrees = sum(degrees, na.rm=T)) %>%
  ungroup %>% 
  filter(!is.na(Discipline)) %>%
  group_by(Discipline, year) %>%
  summarize(degrees = sum(degrees, na.rm=T)) %>%
  ungroup %>% 
  inner_join(totals) %>% 
  mutate(percent = degrees/year_total) %>%
  group_by(Discipline) %>% 
  filter(2007 %in% year) %>%
  mutate(relative = percent/percent[year==2007]) %>%
  filter(year > 1998) %>%
  group_by(Discipline) %>%
  mutate(size = factor(round(.5 + log(mean(percent), base = 9))))
#joint_series %>% count(Discipline, wt = degrees) %>% arrange(-n)
plot2 = long_term2 %>% mutate(source = "old") %>% group_by(Discipline, year) %>%
  summarize(degrees = sum(degrees)) %>% #select(year = Year, Discipline = variable, degrees = count, source) %>% 
  filter(Discipline != "All", year < 2000) %>% 
  bind_rows(
    joint_series %>% 
      #filter(grepl("(English Language and Literature|History|Languages|Philosophy)", Discipline)) %>% 
      filter(!grepl("(Liberal Studies|Communication)", Discipline)) %>%
      select(year, Discipline, degrees) %>% mutate(source="new") %>% filter(year >= 2000)
  ) %>% 
  ungroup %>%
  mutate(Discipline = fct_recode(Discipline, c("English" = "English Language and Literature"), c("Languages" = "Languages and Literatures Other than English"))) %>%
  mutate(Discipline = fct_other(Discipline, keep=c("English", "History", "Languages", "Communication", "Philosophy", "Study of the Arts", "Religion"), other_level = "Misc. fields")) %>%
  group_by(Discipline, year) %>%
  summarize(degrees = sum(degrees)) %>%
  inner_join(pop) %>%
  inner_join(tot_degrees) %>%
  group_by(Discipline) %>% 
  mutate(years = n(), alldeg = sum(degrees)) %>%
  ungroup %>% 
  mutate(Discipline = fct_reorder(Discipline, alldeg) %>% fct_reorder(years)) %>%
  distinct(year, Discipline, .keep_all = T)
# plot2 %>% filter(Discipline=="History")
plot2 %>% ggplot() + 
  geom_area(aes( x = year, y = degrees/collegepop,
                fill=Discipline)) + labs(
    fill = "Discipline",
    y = "Degrees per thousand",
    title = "Degrees per 1000 23-year-olds in the United States",
    subtitle = "Humanities Fields: long term data only for four largest.",
    caption = "Ben Schmidt: Data from IPEDS/Taxonomy from American Academy.\nNot including communication or general Humanities/Liberal Studies"
) + scale_fill_brewer(palette = "Set2", type="qual") + scale_y_continuous() + theme_bw()
```

Perhaps most alarming is that the recent decline is especially severe at liberal arts colleges and more elite schools. While writers often worry too much about the most prestigious universities, they have always been one of the only places where the humanities were central to the mission of higher education.
Elite liberal arts colleges have historically been about evenly divided between the sciences, social sciences, and humanities. But
in the last decade, their humanities majors have fallen from a third to well under a quarter of all degrees. Elite research universities, too, have seen a drop of about to about 70% of their pre-crisis numbers.
The rare schools *not* to have a seen a drop in humanities enrollments tend to be regional comprehensive universities that never had much of a specialty in the humanities to begin with.

```{r}
top30LA = characteristics %>% filter(grepl("^(Williams|Amherst|Bowdoin|Swarthmore|Wellesley|Middlebury|Pomona|Carleton|Claremont McKenna|Davidson|Washington and Lee|Colby|Colgate|Harvey Mudd|Smith|Vassar|Grinnell|Hamilton|Haverford|Wesleyan University|Bates|Colorado College|University of Richmond|Oberlin College|Scripps College|Bryn Mawr)", INSTNM), CARNEGIE==31) %>% mutate(elite = TRUE) %>% select(UNITID, elite)

top30 = characteristics %>% 
  filter(grepl("Harvard|Yale|Princeton|Dartmouth|University of Pennsylvania|Brown|University of Chicago|Stanford|Columbia University|Duke|Massachusetts Institute|University of Pennsylvania|Johns Hopkins|California Institute|Northwestern|Cornell|Rice|Notre Dame|Vanderbilt|Washington University|Georgetown|Berkeley|Southern California|Carnegie Mellon|Los Angeles|University of Virginia|Tufts|Ann Arbor|Wake Forest|Chapel Hill",INSTNM), CARNEGIE==15) %>% 
  mutate(elite = TRUE) %>% 
  select(UNITID, elite)


charelite = top30LA %>% rbind(top30) %>% right_join(characteristics)# %>% left_join(HSIs)

charelite = charelite %>% mutate(
  label = 
    case_when(
      elite ~ case_when(
        CARNEGIE == 15 ~ "Elite Research Universities",
        CARNEGIE == 31 ~ "Elite Liberal Arts Colleges"
      ),
#      HBCU == 1 ~ "HBCU",
      CARNEGIE %in% c(15, 16) ~ "Research Universities",
      CARNEGIE %in% c(21, 22) ~ "Masters College/University",
      CARNEGIE %in% c(31, 32) ~ "Bachelor's Colleges",
      TRUE ~ "Other"
    )
)

d4 = data %>% filter(AWLEVEL==5, MAJORNUM == 1) %>% 
  group_by(year, CIPCODE, UNITID) %>% 
  summarize(degrees = sum(degrees)) %>%
  left_join(amadjusted)

totals = d4 %>% 
  group_by(year, UNITID) %>% 
  summarize(year_total = sum(degrees, na.rm=T), people_total = sum(degrees, na.rm=T)) %>%
  inner_join(charelite) %>%
  group_by(label, year) %>% summarize(year_total = sum(year_total, na.rm=T))


elite_humanities = d4 %>% 
  filter(Bachelors=="Humanities") %>% 
  group_by(UNITID, year, Discipline) %>% 
  summarize(degrees = sum(degrees)) %>%
  inner_join(charelite) %>% 
  group_by(year, label) %>%
  filter(Discipline != "Communication", Discipline != "General Humanities/Liberal Studies") %>%
  summarize(degrees = sum(degrees, na.rm = T)) %>%
  inner_join(totals) %>%
  filter(label != "Other", year > 1997) %>%
  group_by(label) %>%
  mutate(percent = (degrees/year_total), relative = (degrees/year_total)/(degrees/year_total)[year==2008])

elite_humanities %>%
  ggplot() + aes(x=year, y = percent, color=label, label = label) + geom_line() + 
  labs(title = "Humanities degree share by institution class", 
       subtitle = "'Elite' schools in the top 30 per US News and World Report",
       caption = "Ben Schmidt, 2018. Data from IPEDS.\n2000 Carnegie classifications. Not including military academies.") + 
  scale_y_continuous("Percent of degrees at institution type", label = scales::percent) + 
  geom_text(data = elite_humanities %>% filter(label != "Other", year == 1998), adj=1,nudge_x = -.1) + expand_limits(y=0, x = 1989) + theme_bw() +  theme(legend.position="none")


```

At these elite colleges, essentially every field of the humanities has fallen in share: even fields like Classics and Area Studies, which grew through the 2000s. The only bright spot is linguistics, the rare field that directly bridges the humanities and the sciences.

```{r}

p3 = charelite %>% filter(elite) %>% left_join(data) %>% filter(AWLEVEL==5, MAJORNUM==1) %>%
  mutate(gender="all") %>%
  group_by(gender) %>% 
  into_major_nums %>%
  acadjoin %>%
    group_by(Bachelors, Discipline, year) %>%
    filter(Bachelors=="Humanities") %>% 
#    filter(Type %in% c("English","History","Languages and Literature","Philosophy")) %>%
    summarize(degrees = sum(degrees), percent = sum(degrees)/year_total[1]) %>%
  filter(year > 1997) %>%
  mutate(relative = 1 - percent/max(percent), peak_year = year[percent==max(percent)][1])


p3 %>% filter(year==2017, Discipline != "Agriculture") %>% 
  ungroup %>%
  mutate(Discipline = paste0(Discipline, " (peak ", peak_year, ")")) %>% arrange(relative) %>% ungroup %>% select(relative, peak_year, Discipline, percent) %>% ggplot() + geom_point(aes(size=percent, x=reorder(Discipline, peak_year), y = -relative)) + coord_flip() + scale_y_continuous("Change from peak", labels = scales::percent,breaks=seq(-.6, 0, .1)) + scale_x_discrete("Discipline") + scale_size_continuous("Percentage\nof degrees\nin 2017", trans="sqrt", labels = scales::percent, breaks = c(.01, .05, .25, 1, 2)/100) + labs(title="Fall at elite colleges\nin share of humanities majors", caption = "Ben Schmidt, 2018: Data from IPEDS, taxonomy from American Academy") + expand_limits(y=0) + theme_bw() 
```

The timing and placement of the drops at a variety of institutions foils many common explanations for why the humanities are falling.
Do you think student are put off by liberal pieties in the classroom? It's difficult to square that argument with the two decades of stability that followed the beginning of the culture wars in the late 1980s; nor has postmodernism completely swept the field at schools like Brigham Young's main campus (down 20%) or Bob Jones University (down 30%).
Do you blame student loan and the cost of college? Then why has the decline been nearly as strong at schools where student debt is almost non-existent, like Princeton University (down 28%) and the College of the Ozarks (down 44%)?

One common explanation *does* line up with the data fairly well, at least in part: that students fled the humanities after the financial crisis because they became more fearful of the job market.  The chart below shows, in individual panels, a few dozen of the most common majors. The fields that have risen in the past decade are almost entirely STEM majors: majors like nursing, engineering, computer science, and biology. Quantitative social sciences like economics and psychology have held steady, while fields in closer proximity to the humanities like political science, sociology, and anthropology have shown declines, especially since 2011.

```{r, fig.width=12, fig.height=12}
source("EDA_functions.R")

majornums = data %>% filter(MAJORNUM==1, AWLEVEL==5) %>% group_by(year, CIPCODE) %>% summarize(degrees = sum(degrees)) %>% acadjoin %>% filter(year > 1998)

majornums %>% group_by(Bachelors) %>% group_by(Bachelors, Discipline, year) %>% summarize(degrees=sum(degrees)) %>% group_by(year) %>% mutate(percent = degrees/sum(degrees)) %>% 
  group_by(Discipline, Bachelors) %>% filter(year > 1998, sum(degrees) > 10000) %>% ggplot() + geom_line(aes(x=year, y = percent, color = Bachelors)) + facet_wrap(~Bachelors + Discipline, scales="free_y" ) + expand_limits(y=0) + scale_y_continuous(labels = scales::percent) + theme_bw()+ theme(legend.position = "none") + labs(title="Degree Share of common majors", subtitle="Data from NCES IPEDS: Taxonomy building on American Academy of Arts and Science.\nBen Schmidt, 2018")


```

This may seem like a roundabout way to state common knowledge: that STEM majors are the only safe bet in the modern economy, and the humanities are dying out because kids no longer have the luxury of a useless major. But there's an extremely important caveat: students aren't fleeing degrees with poor job prospects. They're fleeing humanities and related fields, specifically, because they *think* they have poor job prospects. If the whole story were a market response to student debt and the great recession, students would have read the 2011 census report calling out [psychology and communications as the fields with the lowest degrees](https://www.census.gov/newsroom/releases/archives/education/cb12-196.html) and fled from them. Or they would have noticed that biology majors make less than the average college graduate, and favored the physical sciences. Most 18-year-olds are not econometricians, and those that are were probably going to major in economics anyway. The census has been asking about college majors for almost a decade now and aside from a few obvious points (engineers make more money than journalists), the results are more surprising for how trivial the differences between majors seem to be. Although the preponderance of evidence indicates that humanities majors are slightest worse off than average. The data--for example, [humanities majors under the age of 35 are less likely to be unemployed than life science or social science majors](https://www.humanitiesindicators.org/content/indicatordoc.aspx?i=10919#fig10922). But the fact is that the differences in earnings and employment the census has turned up are mild. Being the *type of person* inclined to view a college major in terms of return on investment will probably make a much bigger difference in your earnings than the actual major does. And other factors, like gender, matter even more: [men with terminal humanities BAs make more money than women in any field but engineering.](https://www.humanitiesindicators.org/content/indicatordoc.aspx?i=64)

But although it would be better if students knew the actual data about majors--or, at least, we had some psychology major jokes to go along with the ones about art historians--the idea that students should choose majors against some idea of the job market 4 years out is often nuts. Not always--nursing remains a solid bet for the risk-averse. But even as humanities degrees have fallen at elite schools, degrees in computer science have climbed much more quickly than at other schools: in the top 30 universities according to US News, there are now about as many degrees awarded in computer science as in History, English, Languages, Philosophy, Religion, Area Studies, and Linguistics put *together*. Even setting aside our fears about what this will do to cocktail conversation down at the yacht club; how will students know when our society is precisely three years away from having the right number of Ivy league graduates who can write the quicksort algorithm on a whiteboard? CS degrees have already tumbled once in response to shifts in the job market, after the dot-com crash; I wouldn't want to promise an 18-year-old today that they'll have a job in 2022.

```{r fig.width=6, fig.height=4}
source("EDA_functions.R")

majornums = charelite %>% filter(elite) %>% left_join(data) %>% filter(MAJORNUM==1, AWLEVEL==5) %>% group_by(year, CIPCODE, CARNEGIE) %>% summarize(degrees = sum(degrees)) %>% acadjoin %>% filter(year > 1998)

m2 = majornums %>% group_by(Bachelors) %>% group_by(Bachelors, Discipline, year, CARNEGIE) %>% summarize(degrees=sum(degrees)) %>% group_by(year, CARNEGIE) %>% mutate(percent = degrees/sum(degrees)) %>% 
  group_by(Discipline, Bachelors, CARNEGIE) %>% filter(year > 1998)

#m2 %>% filter(CARNEGIE==31) %>% group_by(Discipline) %>% tally(degrees)

m2 %>% filter(CARNEGIE==15) %>% ungroup %>% mutate(Discipline = case_when(
  Bachelors=="Humanities" ~ "All Humanities fields,\nincl. Communications",
  grepl("Computer", Discipline) ~ "Computer Science/\nElectrical Engineering",
  TRUE ~ "Nne"
)) %>% filter(Discipline!="Nne") %>%
  group_by(Discipline, year, CARNEGIE) %>%
  summarize(share = sum(percent, na.rm=T)) %>% ggplot() + 
  labs(title="Computer Science and Humanities degrees at elite institutions", caption="Data from IPEDS: using the top 30\nresearch universities per US News & World Report")+ 
  geom_line(aes(x=year, y = share, color=Discipline)) + expand_limits(y=0) + scale_y_continuous(labels = scales::percent) + theme_bw() #+ theme(legend.position = "none") + labs(title="")

```

So have students really changed what they *want* to major in, or what they think they *should* major in? Suppose college tuition was free and every first-year had a guaranteed job lined up for after graduation. This parallel universe does exist at the military service academies: and in West Point, Annapolis, and Colorado Springs humanities majors are at about the same level as in 2008. There also, interestingly,  doesn't seem to be an extraordinary change in what students say they want to get out of a college education. 
Decades of studies of college freshman have asked incoming students how important various (and nonexclusive) life goals are to them in the context of their college decisions. The drop in humanities majors in the 1970s, [Dennis Ahlburg and Evan Roberts show in a soon-to-be-published book](https://www.routledge.com/The-Changing-Face-of-Higher-Education-Is-There-an-International-Crisis/Ahlburg/p/book/9781138244832), corresponded to a great inversion: in 1970, 7 in 10 students thought it was very important or essential to "develop a meaningful philosophy of life" through education while about 4 in 10 (and 5 in 10 men) put a priority on using it to "make more money"; by the mid-80s, these ratios had flipped. Of all the statistics on the humanities I've seen, I find this one the most depressing--for the past forty years, the percentage of first-year college students who think highly enough of crafting a life philosophy in the course of their studies to muster the energy to fill out a bubble indicating as much has flatlined below one in two: it's little wonder that so few major in the liberal arts in the end.
But although Alburgh and Roberts find a slight tick down in their time series since 2008 (offsetting a post-2001 rise in students seeking the mysteries of life), the current numbers are safely within the range of the last decades.

A few signs suggest the decline might, eventually, end as quickly as it came on. [There are scattered stories of a return to history classes in the age of Trump](https://qz.com/1000017/donald-trump-saves-college-history/) (although nationwide enrollment numbers don't yet bear it out). Thanks to an anti-affirmative action lawsuit against Harvard, that school released information on its applicant pool, 10,000 or so students who imperfectly track the students that will end up scattered across liberal arts colleges and universities. It showed that in the class of 2019--*maybe*--the rush out of humanities disciplines has slowed for the first time in years. And the speed at which humanities majors are dropping has decreased in the last few years from the 2011 to 2014 freefall. 

The humanities, in other words, will almost certainly survive, even if they don't return to being true peers of the social sciences and sciences in American higher learning. To admit the humanities are in crisis doesn't mean conceding that they are being driven extinct. It means, instead, that their place is diminishing, changing both them and the university as whole. The decisions and rhetoric around the humanities matter more now than in--say--the 1990s, as journals, libraries, and universities have to make new sets of decisions around what shape the new humanities will take.

What comes next will be different. The humanities of the boom years in the 60s circled around a tightly constrained common core of English and history. At their best they helped to sustain, recreate, and improve a shared culture that enriched American life; at their worst, they served as a conduit for carefully controlled cultural capital, and ensured that whole classes of people would see that culture as not being for people like them. These fields have not completely abandoned the canon or world history (yes, [colleges still teach Shakespeare](http://doctorcleveland.blogspot.com/2015/04/no-colleges-still-teach-shakespeare.html)), but few would ever claim they serve as stewards of American Civilization.

While history, English, and the rest have faded, only one set of humanities fields without a foot in the sciences has clearly held its own: the much newer (and smaller) disciplines the statistical agency joins together as ethnic, gender, and cultural studies. (It is possible that media studies has also retained share or grown, but hard to tell from the data.) Relatedly, I’ve only found one large class of schools where humanities enrollments have held steady: historically black colleges and universities. HBCUs are also the only institutional class where a majority of students say they’re dedicated to crafting a philosophy of life.

That is: even as the command of culture becomes less central at elite locations, some humanities may be demonstrating their usefulness than ever to students who seek to better understand culture from outside the dominant perspective. The question is how much space any of the humanities can ultimately take up in a university when the dominant perspective continues to warn students away.  