---
title: "Atlantic"
author: "Ben Schmidt"
date: "August 8, 2018"
output: html_document
---

The crisis of the humanities has been around so long that it's almost definitional.
Coverage of individual academic
disciplines like musicology, history, or comparative literature often deals with actual research in the field: but
talk about the humanities *always* seems to focus on their imminent extinction.
In 2010, Wayne Bivens-Tatum provided a [useful walk through the first 50 years of the humanities crisis, until about 1980](https://blogs.princeton.edu/librarian/2010/11/the_crisis_in_the_humanities/).


```{r}

source("read_cips.R")
source("EDA_functions.R")

characteristics = read_csv("hd2017.csv")
cips = return_cips()
data = return_data()
data = data %>%
  filter(CIPCODE != "99.0000", CIPCODE != "99") %>%
  mutate(AWLEVEL=as.numeric(AWLEVEL),MAJORNUM=as.numeric(MAJORNUM))

totals = data %>%
  filter(AWLEVEL==5) %>%
  group_by(year) %>% 
  summarize(year_total = sum(degrees, na.rm=T), people_total = sum(degrees[MAJORNUM==1], na.rm=T))

long_term = read_tsv("long_term.txt")
ltotals = totals %>% group_by(year) %>% summarize(year_total = sum(year_total)) %>% mutate(source="IPEDS") %>% bind_rows(
  long_term %>6% mutate(source="Old", year_total = total) %>% select(year=Year, year_total, people_total=total, source )
  )

pop = read_csv("pop_estimates.csv") %>% select(-junk)

tot_degrees = ltotals %>% group_by(year) %>% summarize(total_degrees = max(year_total))

```


You can be 

[expanded college credits for high school work](https://www.chronicle.com/article/How-the-Crisis-of-the/234185)



[Peter Mandler reports from Great Britain that the humanities are in considerably better shape](https://aeon.co/essays/the-humanities-are-booming-only-the-professors-can-t-see-it).

But something different has been happening with the humanities in the
last ten years: student majors have dropped, rapidly, at a variety of types of institutions. 
This reflects something 

The biggest impediment to understanding the current state of the
humanities is that we already think we know what's going on.

## The crisis of the humanities is recent

Historically, the four largest fields in the humanities have been
English (always the largest), history (which hardly ever identifies as
a social science anymore), a variety of foreign languages and
literature, and philosophy.


```{r}

source("read_cips.R")
source("EDA_functions.R")

characteristics = read_csv("hd2017.csv")
cips = return_cips()
data = return_data()
data = data %>%
  filter(CIPCODE != "99.0000", CIPCODE != "99") %>%
  mutate(AWLEVEL=as.numeric(AWLEVEL),MAJORNUM=as.numeric(MAJORNUM))

totals = data %>%
  filter(AWLEVEL==5) %>%
  group_by(year) %>% 
  summarize(year_total = sum(degrees, na.rm=T), people_total = sum(degrees[MAJORNUM==1], na.rm=T))

long_term = read_tsv("long_term.txt")
ltotals = totals %>% group_by(year) %>% summarize(year_total = sum(year_total)) %>% mutate(source="IPEDS") %>% bind_rows(
  long_term %>6% mutate(source="Old", year_total = total) %>% select(year=Year, year_total, people_total=total, source )
  )

pop = read_csv("pop_estimates.csv") %>% select(-junk)

tot_degrees = ltotals %>% group_by(year) %>% summarize(total_degrees = max(year_total))

```

## It's really a crisis of the qualitative liberal arts.

The term 'liberal arts' used to denote a split between abstract
knowledge and pre-professional.



What's causing 

There has been a book about the humanities crisis every every decade

## It's not about 

## It's not just about employment

## Some stories about why the humanities collapsed




The Trump administration recently announced it would release guidelines about college majors.




