---
title: "Humanities majors yikes"
author: "Ben Schmidt"
date: "August 8, 2018"
output:
  word_document: default
  html_document: default
---
People have been proclaiming the imminent extinction of the humanities
for decades. [A bestselling volume in
1964](https://catalog.hathitrust.org/Record/001450049) warned that a
science-focused world left no room for humane pursuits, even as baby
boomers began to flood the English and history departments of new
universities. Allan Bloom warned about academics putting liberal
ideology before scholarship in 1987; humanities degrees quickly rose.
While coverage of individual academic disciplines like musicology,
history, or comparative literature often deals with the substance of
scholarship, talk of the humanities in general *always* seems to focus
on their imminent extinction. In 2010, Wayne Bivens-Tatum provided a
[useful walk through the first 50 years of the humanities crisis, until
about
1980](https://blogs.princeton.edu/librarian/2010/11/the_crisis_in_the_humanities/).

Because of this long history, I've always been skeptical of claims that
the humanities are in retreat.

But something different has been happening with the humanities since the
2008 financial crisis. Five years ago, I argued that the humanities were
[still near long-term norms in their number of
majors.](http://www.chronicle.com/blognetwork/edgeofthewest/2013/06/10/the-humanities-crisis/)
But since then, I've been watching the [numbers from the Department of
Education](https://nces.ed.gov/ipeds/), and every year things look
worse. Almost every humanities field has seen a rapid drop in majors:
History is down about 45 percent from its 2007 peak, while the number of
English majors has fallen by nearly half since the late 1990s. Student
majors have dropped, rapidly, at a variety of types of institutions.
Declines have hit almost every field in the humanities (with one
interesting exception) and related social sciences, they have not
stabilized with the economic recovery, and they appear to reflect a new
set of student priorities, which are being formed even before they see
the inside of a college classroom.

```{r, include = FALSE}
knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE, cache = TRUE, fig.width = 8, fig.height = 4)
```

```{r, cache = TRUE}
# Preparatory functions to get data.
source("read_cips.R")

characteristics = read_csv("hd2017.csv")
cips = return_cips()
data = return_data()

source("EDA_functions.R")

data = data %>%
  filter(CIPCODE != "99.0000", CIPCODE != "99", !is.na(degrees)) %>%
  mutate(AWLEVEL=as.numeric(AWLEVEL),MAJORNUM=as.numeric(MAJORNUM))

totals = data %>%
  filter(AWLEVEL==5) %>%
  group_by(year) %>% 
  summarize(year_total = sum(degrees, na.rm=T), people_total = sum(degrees[MAJORNUM==1], na.rm=T))

long_term = read_tsv("long_term.txt")
long_term2 = read_csv("long_term_gender.csv") %>% gather(variable, degrees, -year) %>% separate(variable, c("gender","Discipline")) %>% filter(year < 2000) %>% mutate(gender = ifelse(gender=="Men", "men", "women"))

gender_totals = long_term2 %>% filter(Discipline=="All") %>% select(-Discipline) %>% spread(gender, degrees) %>% 
  bind_rows(data %>% filter(AWLEVEL==5, MAJORNUM==1, year > 1999) %>% group_by(year, gender) %>% summarize(degrees = sum(degrees)) %>%  ungroup %>% spread(gender, degrees))

ltotals = totals %>% group_by(year) %>% summarize(year_total = sum(year_total)) %>% mutate(source="IPEDS") %>% bind_rows(
  long_term %>% mutate(source="Old", year_total = total) %>% select(year=Year, year_total, people_total=total, source )
  ) %>% left_join(gender_totals)

pop = read_csv("pop_estimates.csv") %>% select(-junk)

tot_degrees = ltotals %>% group_by(year) %>% summarize(total_degrees = max(year_total))

```

```{r, fig.height=6, fig.width=7}
plottable = data %>% filter(MAJORNUM==1, AWLEVEL==5) %>% 
  into_major_nums %>% 
  left_join(cips,by=c("CIPCODE"="CIPCode"))

this_plot = plottable %>% filter(year >= 1997) %>%
    group_by(year,General,Type) %>%
    filter(General %in% c("Humanities")) %>% 
    filter(Type %in% c("English","History","Languages and Literature","Philosophy", "Classical Studies","Religion")) %>%
    summarize(degrees = sum(degrees), percent = sum(degrees)/year_total[1]) %>% filter(!(Type=="Area Studies" && year < 1997)) %>%
    group_by(Type) %>%
    mutate(relative=percent/max(percent)) %>% mutate(relative = relative - 1) # change in percent

fig1 = this_plot %>% ggplot(aes(x=year,y=relative,fill=Type,lty=Type,label=Type,color=Type)) +
    geom_line() + 
    scale_y_continuous("Share relative to peak", labels=scales::percent) + 
#    facet_wrap(~Type,scales="free_y") + 
   theme(legend.position="none") + 
    labs(title="The big humanities majors\nhave dropped rapidly in the last decade to decline",subtitle="As percentage of all BAs; all US institutions",
         caption = "IPEDS data (preliminary for 2017)\nChart by Ben Schmidt, 2018") + 
  geom_point(data = this_plot %>% filter(relative==1)) + 
  geom_text(data=this_plot %>% filter(year==2017), adj=0, nudge_x = .1) + theme_bw() + theme(legend.position = "none") + expand_limits(x=2026)

fig1$data %>% write_csv(path="outputs/fig1.csv")
fig1
```

One thing I learned earning a history degree is that people usually
announce a "crisis" so they can trot out solutions they came up with
years earlier. I don't have any right now. But the drop in majors since
2008 has been so intense that I now think there is, in the only
meaningful sense of the word, a crisis. We are in a momentum of rapid
change. The decisions we make now will be especially important and will
have continuing ramifications for what American universities look like
for years to come.

Right now, the biggest impediment to thinking about the future of the
humanities is that, thanks to this entrenched narrative of
decline---because we've been crying wolf for so long---we already think
we know what's going on. Because people care about the humanities and
the university, a familiar set of suspects---skyrocketing student debt,
postmodern relativism, disappearing jobs---are once again being trotted
out. But the data suggest something different. In the wake of the 2008
economic meltdown, American students seem to have adopted new views
about what they *should* be studying, in an ill-informed effort to
guarantee success after graduation. And something essential is being
lost in the process.

### The two crises of the humanities

The most reliable indicators about the humanities in American colleges
are reports that all colleges and universities make to the Department of
Education. These run back to about 1950. Since then, the humanities have
seen three eras. The first ran from 1955 to 1985. As normal schools
around the country, set up to educate teachers, transformed into
comprehensive universities, men and women alike poured into English and
History majors; then, when the economy soured and the growth of higher
education slowed in the 1970s, the boom turned to bust, and humanities
majors collapsed nationwide. The second phase began around 1985 and ran
to 2008. This was a long period of stability; majors in the four largest
(and easiest to track over the long term) humanities majors held steady,
with modest fluctuations. Since 2008, the crisis of the humanities has
resumed, with percentage drops that are beginning to approach those of
40 years ago. Unlike the drops of the 70s, though, there's no
pre-existing bubble to deflate. Nor is there a compelling demographic
explanation. Five years ago, it was reasonable to look at these numbers
and conclude [the long-term story is all about
gender](https://www.theatlantic.com/education/archive/2013/12/the-real-reason-the-humanities-are-in-crisis/282441/).
Men majored in humanities fields at the same rate in the 1990s as they
had in the 1950s, while women, seeing more options in the workforce,
increasingly turned to majors in business fields. But the drops since
the financial crisis can be seen among men and women, across racial
groups, and in a wide variety of universities.

```{r}
majornums = data %>% 
  filter(AWLEVEL==5) %>%
  group_by(year, CIPCODE, MAJORNUM, gender) %>%
  summarize(degrees = sum(degrees)) %>%
  acadjoin()

new_series = majornums %>%
  group_by(year, Discipline, gender) %>%
  filter(Bachelors=="Humanities", MAJORNUM==1) %>%
  summarize(degrees = sum(degrees, na.rm=T)) %>%
  ungroup %>% 
  filter(!is.na(Discipline)) %>%
  group_by(Discipline, year, gender) %>%
  summarize(degrees = sum(degrees, na.rm=T)) %>%
  ungroup %>% 
  inner_join(totals) %>% 
  mutate(percent = degrees/year_total) %>%
  group_by(Discipline) %>% 
  group_by(Discipline) %>%
  mutate(size = factor(round(.5 + log(mean(percent), base = 9))))

joint_series = long_term2 %>% mutate(source = "old") %>% select(year, Discipline, gender, degrees = degrees, source) %>% 
  filter(Discipline != "All", year < 2000) %>% 
  bind_rows(
    new_series %>% 
      #filter(grepl("(English Language and Literature|History|Languages|Philosophy)", Discipline)) %>% 
      filter(!grepl("Liberal Studies)", Discipline)) %>%
      select(year, Discipline, degrees, gender) %>% mutate(source="new") %>% filter(year >= 2000)
  ) %>% 
  mutate(Discipline = fct_recode(Discipline, c("English" = "English Language and Literature"), c("Languages" = "Languages and Literatures Other than English"))) %>%
  inner_join(pop) %>%
  inner_join(tot_degrees) %>%
  group_by(Discipline, gender) %>% 
  mutate(years = sum(degrees)) %>%
  mutate(longterm = years > 300000) %>%
  distinct(year, Discipline, gender, .keep_all = T)

```

```{r}
this_plot = joint_series %>% filter(Discipline %in% c("English", "History", "Languages", "Philosophy")) %>% group_by(year, total_degrees, gender) %>% summarize(humanities=sum(degrees)) %>% filter(year >= 1950) %>% inner_join(gender_totals %>% gather("gender", "total", men, women))

#this_plot %>% ggplot() + geom_text(data = this_plot %>% filter(year %in% c(1955, 1960, 1964, 1967, 1969, 1971, 1973, 1977, 1984, 1993, 2000, 2006, 2013, 2017))) + geom_path() + aes(x=total, y = humanities, label=year, color=gender) + expand_limits(y = 0, x = 0) + scale_y_continuous("Degrees in English, Philosophy, History, Languages", labels = scales::comma) + scale_x_continuous("Overall BA degrees", labels = scales::comma)

labels = frame_data(
  ~label, ~gender, ~percent, ~year,
  "Women", "women", .2, 1978,
  "Men", "men", .05, 1977
)

fig2 = this_plot %>% mutate(percent = humanities/total) %>% ggplot() + aes(x=year,  y = percent, color=gender) + geom_line(lwd=1.2) + theme_bw() + scale_y_continuous("History, Philosophy, English, and Languages\nas percent of all degrees", labels=scales::percent) + expand_limits(y=0) + theme(legend.position="none") + labs(title="Long-term Humanities Degree share in American Universities", subtitle="Four largest degree fields") + scale_color_brewer(palette = "Accent", type="qual") + geom_text(data = labels, size=5, aes(label = label))

fig2$data %>% write_csv("outputs/fig2.csv")

fig2
```

College degrees are a somewhat problematic metric: I'd rather see
information about the type and level of courses that undergraduates
take. This kind of data is hard to come by except at an anecdotal level.
But where large scale enrollment numbers do exist, they are not
especially reassuring: The American Historical Association surveyed
history departments and found [course enrollments dropping by 7.7 from
2014 to
2017](https://www.historians.org/publications-and-directories/perspectives-on-history/february-2018/enrollment-declines-continue-aha-survey-again-shows-fewer-undergraduates-in-history-courses),
with upper-level courses seeing greater declines than introductory ones.
This is bad, but not as catastrophic as the numbers of history majors,
which fell by 20 percent in the same period. That suggests that either
the declines are beginning to stabilize, or that students are more
willing to spend elective credits on humanities courses than to major in
them.

There's another way that calculating the humanities' share of all
four-year degrees is a bit strange. American higher education
continuously expands to include new populations; perhaps the
democratization of access has diluted the prominence of the humanities,
even without eroding the absolute number of degrees . And the humanities
are multiplying. There weren't many Chicano Studies majors in 1968;
perhaps students are simply shifting from traditional humanistic fields
of study to new ones.

But even accounting for these factors, the humanities are in retreat.
Looking at degrees as a share of the entire American 23-year-old
population shows the humanities faring better at the turn of the
century. But even after adding a raft of fields like ethnic and gender
studies, musicology, art history, and religion, a lower share of
newly-graduated Americans earn humanities degrees today than did so in
1970 or 1990. Even the absolute *number* is lower than in 1970: The big
four humanities fields---philosophy, history, languages, and
English---are at risk of dipping below 100,000 degrees for the first
time in almost 20 years. (If you include mass communication and media
studies, which mixes social science, humanities, and professional
education together, things look a bit better, but increases in
communication majors since 2008 nationwide can only offset a single
year's loss of English BAs.)

```{r}
library(ggrepel)
joint_series = majornums %>% filter(MAJORNUM==1) %>%
  group_by(year, Discipline) %>%
  filter(Bachelors=="Humanities") %>%
  summarize(degrees = sum(degrees, na.rm=T)) %>%
  ungroup %>% 
  filter(!is.na(Discipline)) %>%
  group_by(Discipline, year) %>%
  summarize(degrees = sum(degrees, na.rm=T)) %>%
  ungroup %>% 
  inner_join(totals) %>% 
  mutate(percent = degrees/year_total) %>%
  group_by(Discipline) %>% 
  filter(2007 %in% year) %>%
  mutate(relative = percent/percent[year==2007]) %>%
  filter(year > 1998) %>%
  group_by(Discipline) %>%
  mutate(size = factor(round(.5 + log(mean(percent), base = 9))))
#joint_series %>% count(Discipline, wt = degrees) %>% arrange(-n)
plot2 = long_term2 %>% mutate(source = "old") %>% group_by(Discipline, year) %>%
  summarize(degrees = sum(degrees)) %>% #select(year = Year, Discipline = variable, degrees = count, source) %>% 
  filter(Discipline != "All", year < 2000) %>% 
  bind_rows(
    joint_series %>% 
      #filter(grepl("(English Language and Literature|History|Languages|Philosophy)", Discipline)) %>% 
      filter(!grepl("(Liberal Studies|Communication)", Discipline)) %>%
      select(year, Discipline, degrees) %>% mutate(source="new") %>% filter(year >= 2000)
  ) %>% 
  ungroup %>%
  mutate(Discipline = fct_recode(Discipline, c("English" = "English Language and Literature"), c("Languages" = "Languages and Literatures Other than English"))) %>%
  mutate(Discipline = fct_other(Discipline, keep=c("English", "History", "Languages", "Communication", "Philosophy", "Study of the Arts", "Religion"), other_level = "Misc. fields")) %>%
  group_by(Discipline, year) %>%
  summarize(degrees = sum(degrees)) %>%
  inner_join(pop) %>%
  inner_join(tot_degrees) %>%
  group_by(Discipline) %>% 
  mutate(years = n(), alldeg = sum(degrees)) %>%
  ungroup %>% 
  mutate(Discipline = fct_reorder(Discipline, alldeg) %>% fct_reorder(years)) %>%
  distinct(year, Discipline, .keep_all = T)
# plot2 %>% filter(Discipline=="History")


fig3 = plot2 %>% mutate(permille = degrees/collegepop) %>% ggplot() + 
  geom_area(aes( x = year, y = permille,
                fill=Discipline)) + labs(
    fill = "Discipline",
    y = "Degrees per thousand",
    title = "Degrees per 1000 23-year-olds in the United States",
    subtitle = "Humanities Fields: long term data only for four largest.",
    caption = "Ben Schmidt: Data from IPEDS/Taxonomy from American Academy.\nNot including communication or general Humanities/Liberal Studies"
) + scale_fill_brewer(palette = "Set2", type="qual") + scale_y_continuous() + theme_bw()

fig3$data %>% write_csv("outputs/fig3.csv")

fig3
```

Perhaps most alarming is that the recent decline has hit
liberal arts colleges and more elite universities--which avoided the worst of the plunge in the 70s--quite hard.
While writers often worry
too much about the most prestigious universities, they have always been
one of the only places where the humanities were central to the mission
of higher education. Elite liberal arts colleges have historically been
about evenly divided between the sciences, social sciences, and
humanities. But in the last decade, their humanities majors have fallen
from a third to well under a quarter of all degrees. Elite research
universities, too, have seen a drop to about 70 percent of their
pre-crisis numbers. The rare schools *not* to have a seen a drop in
humanities enrollments tend to be regional comprehensive universities
that never had much of a specialty in the humanities to begin with.

```{r}
top30LA = characteristics %>% filter(grepl("^(Williams|Amherst|Bowdoin|Swarthmore|Wellesley|Middlebury|Pomona|Carleton|Claremont McKenna|Davidson|Washington and Lee|Colby|Colgate|Harvey Mudd|Smith|Vassar|Grinnell|Hamilton|Haverford|Wesleyan University|Bates|Colorado College|University of Richmond|Oberlin College|Scripps College|Bryn Mawr)", INSTNM), CARNEGIE==31) %>% mutate(elite = TRUE) %>% select(UNITID, elite)

top30 = characteristics %>% 
  filter(grepl("Harvard|Yale|Princeton|Dartmouth|University of Pennsylvania|Brown|University of Chicago|Stanford|Columbia University|Duke|Massachusetts Institute|University of Pennsylvania|Johns Hopkins|California Institute|Northwestern|Cornell|Rice|Notre Dame|Vanderbilt|Washington University|Georgetown|Berkeley|Southern California|Carnegie Mellon|Los Angeles|University of Virginia|Tufts|Ann Arbor|Wake Forest|Chapel Hill",INSTNM), CARNEGIE==15) %>% 
  mutate(elite = TRUE) %>% 
  select(UNITID, elite)


charelite = top30LA %>% rbind(top30) %>% right_join(characteristics)# %>% left_join(HSIs)

charelite = charelite %>% mutate(
  label = 
    case_when(
      elite ~ case_when(
        CARNEGIE == 15 ~ "Elite Research Universities",
        CARNEGIE == 31 ~ "Elite Liberal Arts Colleges"
      ),
      HBCU == 1 ~ "HBCU",
      CARNEGIE %in% c(15, 16) ~ "Research Universities",
      CARNEGIE %in% c(21, 22) ~ "Masters College/University",
      CARNEGIE %in% c(31, 32) ~ "Bachelor's Colleges",
      TRUE ~ "Other"
    )
)

d4 = data %>% filter(AWLEVEL==5, MAJORNUM == 1) %>% 
  group_by(year, CIPCODE, UNITID) %>% 
  summarize(degrees = sum(degrees)) %>%
  left_join(amadjusted)

totals = d4 %>% 
  group_by(year, UNITID) %>% 
  summarize(year_total = sum(degrees, na.rm=T), people_total = sum(degrees, na.rm=T)) %>%
  inner_join(charelite) %>%
  group_by(label, year) %>% summarize(year_total = sum(year_total, na.rm=T))


elite_humanities = d4 %>% 
  filter(Bachelors=="Humanities") %>% 
  group_by(UNITID, year, Discipline) %>% 
  summarize(degrees = sum(degrees)) %>%
  inner_join(charelite) %>% 
  group_by(year, label) %>%
  filter(Discipline != "Communication", Discipline != "General Humanities/Liberal Studies") %>%
  summarize(degrees = sum(degrees, na.rm = T)) %>%
  inner_join(totals) %>%
  filter(label != "Other", year > 1997) %>%
  group_by(label) %>%
  mutate(percent = (degrees/year_total), relative = (degrees/year_total)/(degrees/year_total)[year==2007])

fig4 = elite_humanities %>%
  ggplot() + aes(x=year, y = percent, color=label, label = label) + geom_line() + 
  labs(title = "Humanities degree share by institution class", 
       subtitle = "'Elite' schools in the top 30 per US News and World Report",
       caption = "Ben Schmidt, 2018. Data from IPEDS.\n2000 Carnegie classifications. Not including military academies.") + 
  scale_y_continuous("Percent of degrees at institution type", label = scales::percent) + 
  geom_text(data = elite_humanities %>% filter(label != "Other", year == 1998), adj=1,nudge_x = -.1) + expand_limits(y=0, x = 1989) + theme_bw() +  theme(legend.position="none")

fig4$data %>% write_csv("outputs/fig4.csv")

fig4
```

At these elite colleges, essentially every field of the humanities has
fallen in share---even fields like classics and area studies, which grew
through the 2000s. The only bright spot is linguistics, the rare field
that directly bridges the humanities and the sciences.

```{r}

p3 = charelite %>% filter(elite) %>% left_join(data) %>% filter(AWLEVEL==5, MAJORNUM==1) %>%
  mutate(gender="all") %>%
  group_by(gender) %>% 
  into_major_nums %>%
  acadjoin %>%
    group_by(Bachelors, Discipline, year) %>%
    filter(Bachelors=="Humanities") %>% 
#    filter(Type %in% c("English","History","Languages and Literature","Philosophy")) %>%
    summarize(degrees = sum(degrees), percent = sum(degrees)/year_total[1]) %>%
  filter(year > 1997) %>%
  mutate(relative = 1 - percent/max(percent), peak_year = year[percent==max(percent)][1])

fig5 = p3 %>% filter(year==2017, Discipline != "Agriculture") %>% 
  ungroup %>%
  mutate(Discipline = paste0(Discipline, " (peak ", peak_year, ")")) %>% arrange(relative) %>% ungroup %>% select(relative, peak_year, Discipline, percent) %>% ggplot() + geom_point(aes(size=percent, x=reorder(Discipline, peak_year), y = -relative)) + coord_flip() + scale_y_continuous("Change from peak", labels = scales::percent,breaks=seq(-.6, 0, .1)) + scale_x_discrete("Discipline") + scale_size_continuous("Percentage\nof degrees\nin 2017", trans="sqrt", labels = scales::percent, breaks = c(.01, .05, .25, 1, 2)/100) + labs(title="Fall at elite colleges\nin share of humanities majors", caption = "Ben Schmidt, 2018: Data from IPEDS, taxonomy from American Academy") + expand_limits(y=0) + theme_bw() 

fig5$data %>% write_csv(path = "outputs/fig5.csv")

fig5

```

The timing and placement of the drops at a variety of institutions foils
many common explanations for why the humanities are falling. Do you
think students are put off by liberal pieties in the classroom? It's
difficult to square that argument with the two decades of stability that
followed the beginning of the culture wars in the late 1980s; nor has
postmodernism completely swept the field at schools like Brigham Young's
main campus (down 20 percent) or Bob Jones University (down 50 percent).
Do you blame student loans and the cost of college? Then why has the
decline been nearly as strong at schools where student debt is almost
non-existent, like Princeton University (down 28 percent) and the
College of the Ozarks (down 44 percent)?

One common explanation *does* line up with the data fairly well, at
least in part: that students fled the humanities after the financial
crisis because they became more fearful of the job market. The chart
below shows, in individual panels, a few dozen of the most common
majors. The fields that have risen in the past decade are almost
entirely STEM majors, including nursing, engineering, computer science,
and biology. Quantitative social sciences like economics and psychology
have held steady, while fields in closer proximity to the humanities
like political science, sociology, and anthropology have shown declines,
especially since 2011.

```{r, fig.width=12, fig.height=12}
source("EDA_functions.R")

majornums = data %>% filter(MAJORNUM==1, AWLEVEL==5) %>% group_by(year, CIPCODE) %>% summarize(degrees = sum(degrees)) %>% acadjoin %>% filter(year > 1998)

fig6 = majornums %>% group_by(Bachelors) %>% group_by(Bachelors, Discipline, year) %>% summarize(degrees=sum(degrees)) %>% group_by(year) %>% mutate(percent = degrees/sum(degrees)) %>% 
  group_by(Discipline, Bachelors) %>% filter(year > 1998, sum(degrees) > 10000) %>% 
  ggplot() + geom_line(aes(x=year, y = percent, color = Bachelors)) + facet_wrap(~Bachelors + Discipline, scales="free_y" ) + expand_limits(y=0) + scale_y_continuous(labels = scales::percent) + theme_bw()+ theme(legend.position = "none") + labs(title="Degree Share of common majors", subtitle="Data from NCES IPEDS: Taxonomy building on American Academy of Arts and Science.\nBen Schmidt, 2018")

fig6$data %>% write_csv(path = "outputs/fig6.csv")

fig6
```

This may seem like a roundabout way to present the conventional wisdom
that STEM majors are the only safe bet in the modern economy, and the
humanities are dying out because kids no longer have the luxury of a
useless major. But there's an extremely important caveat: Students
aren't fleeing degrees with poor job prospects. They're fleeing
humanities and related fields, specifically, because they *think* they
have poor job prospects. If the whole story were a market response to
student debt and the great recession, students would have [read the 2011
census report naming communication and psychology among the lowest-earning degrees](https://www.census.gov/newsroom/releases/archives/education/cb12-196.html)
and fled from them. Or they would have noticed that
biology majors make less than the average college graduate, and favored
the physical sciences.

For almost a decade, the census has been regularly asking a random sample of Americans about their college major.
This gives us far rich information about career
outcomes than the lists of starting salaries that occasionally
circulate. Some commonly-held wisdom is true--engineers make more money
than any other field---but the results are most surprising for how trivial the
differences between most majors turn out to be. Humanities majors
[have the highest medical school acceptance rates of any field](https://www.aamc.org/download/321496/data/factstablea17.pdf), 
show [lower rates of long-term underemployment that most fields](https://www.myajc.com/blog/get-schooled/opinion-you-can-make-more-than-lattes-with-that-humanities-degree/p63EL6SfM2o6DLd6exVnRM/),
and [humanities majors under the age of 35 are *less* likely to be unemployed than life-science or social-science
majors.](https://www.humanitiesindicators.org/content/indicatordoc.aspx?i=10919#fig10922)

Still, the [balance of evidence](https://nces.ed.gov/programs/coe/pdf/coe_sbc.pdf) suggests that humanities majors are probably
slightly worse off than average---maybe as much as one more point of
unemployment and \$5,000 to \$10,000 a year in income. They aren't alone, of course: while finance and
computer science majors make more, biology and business majors make
about the same, and psychology majors usually come out worse.
Other factors, like gender, matter more: [Men with terminal humanities
BAs make more money than women in any field but
engineering.](https://www.humanitiesindicators.org/content/indicatordoc.aspx?i=64)
Being the *type of person* inclined to view a college major in terms of
return on investment will probably make a much bigger difference in your
earnings than the actual major does.

In other areas of the economy, we view these kinds of differences with
equanimity. The difference between humanities majors and science majors,
in median income and unemployment, is roughly the same as the difference between
residents of Virginia and North Carolina. If someone told to me not to
move to Charlotte because no one there makes a living, I would never
take them seriously. But worried relatives express the same concerns
about classics majors every day, with no sounder evidence.

It would be better if students knew the actual data about majors---or,
at least, if there were some psychology-major jokes to go along with the
ones about art historians. But the idea that students should choose
majors by trying to guess what the job market will reward several years
later is often nuts. There are some consistent trends; nursing remains a
solid bet for the risk-averse. Other majors, though, may prove greater
gambles. As humanities degrees have fallen at elite schools, degrees in
computer science have climbed much more quickly than at other schools.
In the top 30 universities, according to *U.S. News*, there are now
about as many degrees awarded in computer science and electrical engineering as in history,
English, languages, philosophy, and area studies
put together. Set aside fears about what this will do to cocktail
conversation down at the yacht club; how are students supposed to know when the
U.S. is precisely three years away from having the right number of Ivy
League graduates who can code onto a whiteboard?
Computer science degrees have already tumbled once in response to shifts
in the job market, after the dot-com crash; I wouldn't want to promise
18-year-olds today that a computer science degree will make them stand
out on the job market in 2022.

```{r fig.width=6, fig.height=4}
source("EDA_functions.R")

majornums = charelite %>% filter(elite) %>% left_join(data) %>% filter(MAJORNUM==1, AWLEVEL==5) %>% 
  group_by(year, CIPCODE, CARNEGIE) %>% summarize(degrees = sum(degrees)) %>% 
  acadjoin %>% filter(year > 1998)

m2 = majornums %>% group_by(Bachelors) %>% group_by(Bachelors, Discipline, year, CARNEGIE) %>% summarize(degrees=sum(degrees)) %>% group_by(year, CARNEGIE) %>% mutate(percent = degrees/sum(degrees)) %>% 
  group_by(Discipline, Bachelors, CARNEGIE) %>% filter(year > 1998)

{m2 %>% filter(CARNEGIE==15) %>% group_by(Discipline, Bachelors) %>% tally(degrees) %>% arrange(-n) %>% filter(Bachelors=="Humanities")}$Discipline

fig7 = m2 %>% filter(CARNEGIE==15) %>% ungroup %>% 
  mutate(Discipline = case_when(
    Discipline=="Communication" ~ "nne",
    Discipline=="Study of the Arts" ~ "nne",
    Discipline %in% c("English Language and Literature" ,             "History", "Languages and Literatures Other than English", "Area Studies", "Philosophy","Comparative Literature") ~ "English, History,\nLanguages and Literature,\nArea Studies and Philosophy",
  grepl("Computer|lectrical", Discipline) ~ "Computer Science/\nElectrical Engineering",
  TRUE ~ "nne"
)) %>% filter(Discipline!="nne") %>%
#  filter(Bachelors=="Humanities" | grepl("Computer|Electrical Engineering", Discipline)) %>%
  group_by(Discipline, year, CARNEGIE) %>%
  summarize(share = sum(percent, na.rm=T)) %>% ggplot() + 
  labs(title="Computer Science and select Humanities degrees at elite universities", caption="Data from IPEDS: using the top 30\nresearch universities per US News & World Report")+ 
  geom_line(aes(x=year, y = share, color=Discipline)) + expand_limits(y=0) + scale_y_continuous(labels = scales::percent) + theme_bw() #+ theme(legend.position = "none") + labs(title="")
fig7

fig7$data %>% write_csv(path = "outputs/fig7.csv")
```

So does the crisis in the humanities actually reflect a shift in what
students want to select as a major, or is it just a change in what they
think they *should* choose as a major? If students legitimately find
computers more interesting than history in the age of Facebook, the fall
of the humanities might not be a bad thing. Suppose college tuition was
free and every first-year had a guaranteed job lined up for after
graduation. This parallel universe does exist at the military service
academies---and at West Point, Annapolis, and Colorado Springs,
humanities majors are at about the same level as they were in 2008.

There also doesn't seem to be an extraordinary change in what students
say they want to get out of a college education. Decades of studies of
college freshman have asked incoming students how important various (and
nonexclusive) life goals are to them in the context of their college
decisions. The drop in humanities majors in the 1970s, [Dennis Ahlburg
and Evan Roberts show in a soon-to-be-published
book](https://www.routledge.com/The-Changing-Face-of-Higher-Education-Is-There-an-International-Crisis/Ahlburg/p/book/9781138244832),
corresponded to a great inversion. In 1970, seven in 10 students thought
it was very important or essential to "develop a meaningful philosophy
of life" through education while about four in 10 (and five in 10 men)
put a priority on using it to "make more money." By the mid-'80s, these
ratios had flipped. Of all the statistics on the humanities I've seen, I
find this one the most depressing---for the past 40 years, the
percentage of first-year college students who think highly enough of
crafting a life philosophy in the course of their studies to muster the
energy to fill out a bubble indicating as much has flatlined below half.
It's little wonder that so few major in the liberal arts in the end. But
although Alburgh and Roberts find a slight tick down in their time
series since 2008 (offsetting a post-2001 rise in students seeking the
mysteries of life), the current numbers are safely within the range of
the last decades.

A few signs suggest the decline might, eventually, end as quickly as it
came on. [There are scattered stories of a return to history classes in
the age of
Trump](https://qz.com/1000017/donald-trump-saves-college-history/)
(although nationwide enrollment numbers don't yet bear it out). Thanks
to an anti-affirmative action lawsuit against Harvard, that school
released information on its applicant pool, 10,000 or so students who
imperfectly track the students who will end up scattered across liberal
arts colleges and universities. It showed that in the class of 2019, the
rush out of humanities disciplines may have slowed for the first time in
years. And the speed at which humanities majors are dropping has
decreased in the last few years, after the freefall between 2011 and
2014.

The humanities, in other words, will almost certainly survive, even if
they don't return to being true peers of the social sciences and
sciences in American higher education. To admit the humanities are in
crisis doesn't mean conceding that they are being driven extinct. It
means, instead, that their place is diminishing, changing both them and
the university as whole. The decisions and rhetoric around the
humanities now have especial importance, as journals, libraries, and
universities have to make new sets of decisions around what shape the
new humanities will take.

What comes next will be different. The humanities of the boom years in
the '60s circled around a tightly constrained common core of English and
history. At their best they helped to sustain, recreate, and improve a
shared culture that enriched American life; at their worst, they served
as a conduit for carefully controlled cultural capital, and ensured that
whole classes of people would see that culture as not being for people
like them. These fields have not completely abandoned the canon (yes,
[colleges still teach
Shakespeare](http://doctorcleveland.blogspot.com/2015/04/no-colleges-still-teach-shakespeare.html)),
but few would still claim they serve as stewards of American
civilization.

While history, English, and the rest have faded, only one set of
humanities fields without a foot in the sciences has clearly held its
own: the much newer (and smaller) disciplines the statistical agency
joins together as ethnic, gender, and cultural studies. (It is possible
that media studies has also retained share or grown, but hard to tell
from the data.) Relatedly, I've only found one large class of schools
where humanities enrollments have held steady: historically black
colleges and universities. HBCUs are also the only institutional class
where a majority of students say they're dedicated to crafting a
philosophy of life.

Even as the command of culture becomes less important to education at elite locations,
some humanities may be demonstrating more usefulness than ever to
students who seek to better understand culture from outside the dominant
perspective. The question is how much space any of the humanities can
ultimately take up in a university, when the dominant perspective
continues to warn students away. 